# Linear Regression in R {#RRegression}
In the Module \@ref(LinearRegression) you learned about the theory underlying linear regression and answered application questions related to the results of a linear regression. In this module, you will learn how to perform a linear regression, visualize the results, and make predictions from a linear regression. You will also be asked to answer the same series of questions you answered in Module \@ref(LinearRegression).

Methods in this module require the `NCStats` package (as always) and the `ggplot2` package (for making graphs).

```{r}
library(NCStats)
library(ggplot2)
```

#### Data Sets {-}
The descriptions below will use the same `weight` (lbs) and highway miles per gallon (`HMPG`) data for 93 cars from the 1993 model year that were used in the second example in the Modules \@ref(LinearRegression). These data are in 93cars.csv ([data](https://raw.githubusercontent.com/droglenc/NCData/master/93cars.csv), [meta](https://raw.githubusercontent.com/droglenc/NCData/master/93cars_meta.txt)) and are loaded into R below with the methods described in Section \@ref(alternative-forms-of-getting-data).^[The `which=` argument in `headtail()` is used to show only a few variables of `cars93`, just to save space.]

```{r eval=FALSE}
cars93 <- read.csv("93cars.csv")
headtail(cars93,which=c("MFG","Model","Type","Weight","HMPG"))
```
```{r echo=FALSE}
cars93 <- read.csv("https://raw.githubusercontent.com/droglenc/NCData/master/93cars.csv")
headtail(cars93,which=c("MFG","Model","Type","Weight","HMPG"))
```

&nbsp;

## Fitted Line Plot
The linear fitted line can be added to a scatterplot (see Section \@ref(scatterplots)) with `geom_smooth()`. The `geom_smooth()` function requires `method="lm"` to show the linear regression line and `se=FALSE` to remove the underlying confidence band.^[This is an advanced topic for a future course.]

```{r}
ggplot(data=cars93,mapping=aes(x=Weight,y=HMPG)) +
  geom_point(pch=21,color="black",fill="darkgray",size=2) +
  labs(x="Weight (lbs)",y="Highway MPG") +
  theme_NCStats() +
  geom_smooth(method="lm",se=FALSE)
```

::: {.tip data-latex=''}
Use `method="lm"` in `geom_smooth()` to add the regression line to an existing scatterplot.
:::

## Fitting the Regression Model in R
The linear regression model is fit to two quantitative variables with `lm()`. The first argument is a formula of the form `response~explanatory`, where `response` contains the quantitative response variable and `explanatory` contains the quantitative explanatory variable. The corresponding data frame is given in `data=`. The results of `lm()` may be assigned to an object so that specific results can be extracted.

```{r}
( lm1 <- lm(HMPG~Weight,data=cars93) )
```

The regression results above show that the intercept is `r formatC(coef(lm1)[1],format="f",digits=1)` and the slope is `r formatC(coef(lm1)[2],format="f",digits=4)`.

::: {.tip data-latex=''}
The slope is under the name of the explanatory variable and the intercept is under "(Intercept)" in the `lm()` results.
:::

## Coefficient of Determination
The coefficient of determination is computed by submitting the saved `lm()` object to `rSquared()`. For example, `r formatC(rSquared(lm1,percent=TRUE),format="f",digits=1)`\% of the variability in highway MPG for a car is explained by knowing its weight. [*Note the use of `digits=` to control the number of decimals.*]

```{r}
rSquared(lm1,digits=3)
```

## Predicted Values
Predicted values from the linear regression are obtained with `predict()`, which requires the saved `lm()` object as its first argument. The second argument is a data frame constructed with `data.frame()` that contains the **EXACT** name of the explanatory variable as it appeared in `lm()` set equal to the value of the explanatory at which the prediction should be made. For example, the predicted highway MPG for a car that weights 2500 lbs is `r formatC(predict(lm1,data.frame(Weight=2500)),format="f",digits=1)`, as obtained below.

```{r}
predict(lm1,data.frame(Weight=2500))
```

&nbsp;
